Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover.

### 2022-08-22 reports/manuscript_v5.Rmd
updated manuscript to v5

---
title: "Rodent trapping studies can improve our understanding of the risk of endemic zoonoses spillover and novel zoonotic pathogen emergence."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
  # pdf_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: harvard.csl
latex_engine: xelatex
---

@@

Rodents are important globally distributed reservoirs of known and novel zoonotic pathogens. Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of endemic zoonoses spillover events. These changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk. However, available species distribution and host-pathogen association datasets such as International Union for Conservation of Nature (IUCN) distribution maps, Global Biodiversity Information Facility (GBIF) species occurrence data, or the CLOVER host-pathogen dataset are biased by the inclusion of limited data sources and spatial biases in sampled locations. Here, we synthesise data from West Africa from 127 rodent trapping studies, conducted between 1964-2021, including over 72,000 individual rodent detections to more comprehensively characterise the range and presence of important zoonotic pathogen host species in this region. We identify that rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and calculate the discrepancies between sampling in GBIF and included studies. For five regionally important zoonotic pathogens (Arenaviridae spp., Borrelia spp., *Lassa mammarenavirus*, Leptospira spp. and *Toxoplasma gondii*) we identify host-pathogen associations that are not reported in the consolidated CLOVER dataset. These host-pathogen asssociation omissions may bias studies investigating the current risk of endemic zoonoses and the effect of land use and climate change on future novel pathogen emergence. Finally, we find that for these five pathogen groups, the proportion of a rodent hosts range that has been sampled remains small. A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk. In the interim, future studies must incorporate a measure of the current sampling biases. The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support a more complete understanding of the hazard of zoonotic spillover events.

@@

There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_nodate]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. The impact of endemic zoonoses meanwhile remains underestimated [@maudlin_neglected_2009]. Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@halliday_endemic_2015; @molyneux_zoonoses_2011; @judson_zoonoses_2021]. West Africa is a nexus for these processes and is identified as a hotspot for increased risk of large epidemics of known zoonotic diseases and novel zoonotic pathogen emergence [@allen_global_2017; @grace_mapping_2012; @han_global_2016]. Two taxonomic groups - rodents (Rodentia) and bats (Chiroptera) - contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015]. Specifically, West Africa has been identified as a region at increased hazard for rodent-borne zoonotic disease spillover events under different projected future land-use change scenarios [@garcia-pena_land-use_2021]. Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020].

Rodent species form diverse assemblages, which provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. The role of rodent species in zoonotic disease spillover or novel zoonotic pathogens emergence are examples of ecosystem disservices. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Rodent species with “fast” life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described sampling biases [@bowler_temporal_nodate; @boakes_distorted_2010]. These sampling biases can gravely distort produced species distribution models [@beck_spatial_2014]. Datasets on host-pathogen associations also suffer from biases introduced from literature selection criteria and taxanomic discrepancies a recently produced consolidated dataset (CLOVER) has been produced to reduce this [@gibb_viralemergenceclover_2021]. Studies identifying potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are typically based on these datasources for modelling host species distributions and host-pathogen associations which can result in biased hazard estimates [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. While other regions remain systematically undersampled (e.g., areas of sparse human populations, endemic zoonoses), potentially resulting in a reduced hazard of zoonotic spillover events or novel zoonotic pathogen emergence being attributed [@beck_spatial_2014]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets and providing additional data to minimise inherent biases [@bovendorp_optimising_2017]. Rodent trapping studies describe rodent population assemblages, their geographic distribution and host-pathogen associations beyond what is currently available in consolidated datasets. Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, individual level data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise rodent trapping studies conducted across West Africa from a search of literature between 1964-2022. First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification. Second, we compare this to curated host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with a consolidated dataset (CLOVER) to explore discrepancies in rodent host-pathogen associations and report the proportion of positive assays for pathogens of interest. Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of sparse sampling of pathogens within their host ranges. We expect that rodent trapping studies provide an important additional source of high-resolution data that can be used to enrich available consolidated datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.

@@

For further analysis, we included studies if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation. We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English. One author screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage; title screening, abstract screening and full text review, a random subset (10%) was reviewed by a second author.

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised]. We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover. To understand potential limitations in the representativeness of rodent trapping we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance. For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained. This was performed by multiplying the total number of trapped individuals within that study site by the median trap success for study sites with the same reported habitat type. Stratification was used as trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared with other habitats (3%, IQR 1-9%)

The method of taxonomic identification applied within a study was recorded, reported species identification was assumed to be accurate. The number of individuals of these species or genera was extracted with species and genus names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021] for internal consistency. We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.

Geographic locations of trapping studies was extracted using GPS locations for the most precise location presented. GPS coordinates for locations not reported were produced using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study. All locations were converted to decimal degrees for internal consistency. The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity. The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1). For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

@@

We summarised the number of studies, the year in which trapping occurred and the country in which they were conducted. We identified 4,692 relevant citations, with 127 rodent trapping studies included (Supplementary Table 2.). The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000. The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.). Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).

Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success. A total of 942,629 trap nights were estimated (Figure 1B.). A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.

In total 76,275 small mammals were trapped with 65,628 (90%) identified to species level and 7,439 (10%)  identified to genus, with the remaining classified to higher taxonomic level. The majority of the 132 identified species were Rodentia (102, 78%), of which Muridae (73, 72%) were the most common family. Soricomorpha were the second most identified order of small mammals (28, 21%). 57 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens. Most studies tested for a single microorganism (48, 84%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (21, 37%), *Borrelia sp.* (9, 16%), *Bartonella sp.* (4, 7%) and *Toxoplasma gondii* (4, 7%). Most studies used Polymerase Chain Reaction (PCR) to detect microorganisms (37, 65%), with fewer studies using serology-based tests (11, 19%) or histological or direct visualisation assays (11, 21%). From 32,920 individual rodent samples we produced 351 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms. Further descriptive information from the included studies including geolocated trapping of species, their detection and non-detection alongside microorganism data is available as supplementary material [@simons_rodents_2022].

@@

To investigate the risk to human health from endemic zoonoses and novel pathogen emergence using synthesised data from rodent trapping data understanding of sampling biases is vital. Sampling of rodents has not been performed randomly across space in West Africa (Figure 1.) and at varying intensity, we aimed quantify this bias as the relationship between trapping effort (measured as trap night density (TN density)), human population density and land use type. We define this as relative trapping effort bias.

To address this question, we calculated TN density within each West African level-2 administrative region, administrative boundaries were obtained from GADM 4.0.4 [@database_of_global_administrative_areas_gadm_2022 of Global Administrative Areas]. We assigned trap nights to specific regions using GPS coordinates obtained from trapping studies as described above. The sf package in the R statistical language was used to handle shapefiles and geographic data [@pebesma_simple_2018]. Density was calculated by dividing the number of trap nights by the area of a level-2 administrative area. For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution [@socioeconomic_data_and_applications_center_gridded_2021].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. The proportion of cropland, shrubland, tree cover and urban land cover within a level-2 administrative region in 2005 was calculated.

We investigated the role of these covariates on the spatial bias of TN density by comparing to a spatial model (equation 1), as follows:

$$

\begin{equation}
\tag{1}
\text{TN density} \sim \text{Tweedie((Longitude * Latitude))}

\tag{2}
\text{TN density} \sim \text{Tweedie((Population density) + (Longitude * Latitude))}

\tag{3}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Longitude * Latitude))}

\text{TN density} \sim \text{Tweedie((Proportion cropland) + (Proportion tree cover) + (Proportion urban) + (Longitude * Latitude))}

\tag{5}
\text{TN density} \sim \text{Tweedie((Population density) + (Region area) + (Proportion urban) + (Longitude * Latitude))}
\end{equation}

$$

We formulated these models as a Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude by latitude) [@pedersen_hierarchical_2019]. These models investigated the association between TN density, proportion of urban, cropland, tree cover and human population density. The models were constructed in the *mgcv* package using the R programming language [@wood_generalized_2017; @r_core_team_r_2020]. Selection of the most parsimonious model was based on Deviance explained and the Akaike information criterion for each model. We performed sensitivity analysis by removing sites with imputed trapping effort and performing a pixel-based analysis.

@@

Curated datasets (i.e., IUCN and GBIF) are commonly used in assessments of biodiversity in response to land use and climate change in addition to being used to explore zoonotic pathogen host distributions. To understand any biases introduced from these sources into these analyses we aimed to assess the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies. We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021]. Distributions were cropped to the study region for globally distributed rodent species. We obtained rodent presence locations from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021]. Presence locations were associated to cells of raster with a ~20km resolution produced for the study region.

We assessed the taxonomic coverage of the trapped taxa and for the seven species with the most trap locations (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *A. niloticus*, *Praomys daltoni* and *Cricetomys gambianus*) we mapped detection and non-detection, associating each location with a ~20km resolution raster.  For these seven species we calculated the area of the IUCN expected range. We calculated the percentage of this range covered by presence detections in GBIF, this process was repeated for presence detections from rodent trapping studies. For both GBIF data and trapping studies we calculated the area of detections outside of the IUCN expected range. For rodent trapping studies only, we were able to calculate the area of non-detections within the IUCN expected area. Finally, we calculated the combined area of detection when combining GBIF and rodent trapping study data.

@@

Studies investigating the risk of endemic zoonoses and novel pathogen emergence generally rely on consolidated datasets of host-pathogen associations. To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen asociations with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021]. CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi. An archived version of the CLOVER dataset was obtained and subset for host species trapped within included studies, spatial filtering of CLOVER for host-pathogen associations limited to the study region was not possible [@gibb_viralemergenceclover_2021].

We produced host-pathogen pairs from rodent trapping studies investigating zoonoses. For each host and pathogen, with both identified to species level, we compared our host-pathogen pair network with one produced using CLOVER, reporting concordance and discordance. We describe these networks and report differences in network composition. For host-pathogen pairs with assay results consistent with acute or prior infection we calculate the proportion positive and identify those absent from CLOVER. We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.

### What is the spatial extent of pathogen testing within a hosts range?

Sampling of zoonotic pathogens may also be biased within their hosts ranges, which may hinder inference across the entire range of the species. To quantify this, we focused on the five pathogen groups and species sampled at the most locations (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*). We geolocated individuals of host species tested for these pathogens and calculated the area sampled. We adopt a similar approach to above by mapping sampled hosts to ~20km pixels and calculating the proportion of their IUCN range in which sampling has occurred. We also present the area covered of all areas in which the host has been detected in this study to produce a measure of relative completeness of sampling within the included rodent trapping studies.

Data and code implemented in the R statistical language has been made available alongside this manuscript to reproduce these analyses [ref_zenodo].

@@

table_2 <- read_rds(here("tables", "Table_2_updated_pixel.rds"))%>%
  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled potential rodent host species. * no IUCN range in West African") %>%
  autofit()

>>>>

---
title: "Rodent trapping studies as an overlooked information source for understanding endemic and novel zoonotic spillover."
date: "`r Sys.Date()`"
author: 
  - "David Simons"
  - "Lauren A. Attfield"
  - "Kate E. Jones"
  - "Deborah Watson-Jones"
  - "Richard Kock"
output: 
  officedown::rdocx_document:
    tables:
      caption:
       style: false
    plots:
      caption:
        style: false
  # pdf_document
always_allow_html: yes
bibliography: citations/manuscript_citations.bib
csl: harvard.csl
latex_engine: xelatex
---

@@

Rodents are important globally distributed reservoirs of known and novel zoonotic pathogens. Ongoing anthropogenic land use change is altering the composition of host species assemblages and modifying the risk of zoonoses spillover events. These changes mean that an understanding of the current distribution of rodent species is vital for accurately describing disease hazard and managing risk. However, available species distribution and host-pathogen association datasets (e.g. IUCN, GBIF, CLOVER) are often taxonomically and spatially biased. Here, we synthesise data from West Africa from 127 rodent trapping studies, published between 1964-2022, as an additional source of information to characterise the range and presence of important zoonotic pathogen host species in this region. We identify that these rodent trapping studies, although biased towards human dominated landscapes across West Africa, can usefully complement current rodent species distribution datasets and we calculate the discrepancies between these datasets. For five regionally important zoonotic pathogens (Arenaviridae spp., Borrelia spp., *Lassa mammarenavirus*, Leptospira spp. and *Toxoplasma gondii*), we identify host-pathogen associations that have not been previously reported in host-association datasets. These omissions have the potential for biasing estimates of current risk and drivers of zoonoses. Finally, for these five pathogen groups, we find that the proportion of a rodent hosts range that has been sampled remains small. A priority of future rodent trapping studies should be to sample rodent hosts across a greater geographic range to better characterise current and future risk. In the interim, future studies of spatial pathogen risk informed by rodent distributions must incorporate a measure of the current sampling biases. The current synthesis of contextually rich rodent trapping data enriches available information from IUCN, GBIF and CLOVER which can support a more complete understanding of the hazard of zoonotic spillover events.

@@

There is increasing awareness of the global health and economic impacts of novel zoonotic pathogen spillover, evidenced by the ongoing SARS-CoV-2 pandemic and previous HIV/AIDs and Spanish Influenza pandemics [@bernstein_costs_2022]. The number of zoonotic disease spillover events and the frequency of the emergence of novel zoonotic pathogens from rodents are predicted to increase under intensifying anthropogenic pressure driven by increased human populations, urbanisation, intensification of agriculture, climate change and wildlife defaunation [@allen_global_2017; @hassell_urbanization_2017; @garcia-pena_land-use_2021; @mcmahon_ecosystem_2018]. The impact of endemic zoonoses meanwhile remains underestimated [@maudlin_neglected_2009]. Endemic zoonoses disproportionally affect those in the poorest sections of society, those living in close contact with their animals and those with limited access to healthcare [@halliday_endemic_2015; @molyneux_zoonoses_2011; @judson_zoonoses_2021].

Rodents along with bats contribute the greatest number of predicted novel zoonotic pathogens and known zoonoses [@han_rodent_2015; @gibb_mammal_2021]. Of 2,220 extant rodent species, 244 (10.7%) are described as reservoirs of 85 zoonotic pathogens [@han_rodent_2015], although many species provide important and beneficial ecosystem services including pest regulation and seed dispersal [@fischer_ecosystem_2018]. Rodents typically demonstrate “fast” life histories [@dobson_fast_2007] with traits such as early maturation and short gestation times (<4 days) further associated with being zoonotic reservoirs [@han_rodent_2015; @albery_fast-lived_2021]. Rodent species with “fast” life histories thrive in human dominated landscapes, displacing species less likely to be reservoirs of zoonotic pathogens [@gibb_zoonotic_2020]. The widespread occurrence of reservoir species and their proximity to human activity make the description of rodent species assemblages and host-pathogen associations vitally important to understanding the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence [@han_global_2016].

Despite the importance of understanding these complex systems, current evidence on host-pathogen associations is considerably affected by taxonomic and geographical sampling biases [@gibb_ecosystem_2020; @gibb_mammal_2021]. Curated biodiversity datasets such as the Global Biodiversity Information Facility (GBIF) and resources produced by the International Union for Conservation of Nature suffer from well described spatial and temporal sampling biases [@boakes_distorted_2010; @bowler_temporal_2022]. These sampling biases can importantly distort produced species distribution models [@beck_spatial_2014]. Datasets on host-pathogen associations also can suffer from biases introduced from literature selection criteria and taxanomic discrepancies. These biases are important because identification of potential geographic hotspots of zoonotic disease spillover and novel pathogen emergence are often based on these types of host species distributions and host-pathogen associations [@plowright_sampling_2019; @carlson_future_2021]. For example, systematically increased sampling, over-representation of certain habitats and clustering around areas of high human population could lead to an apparent association between locations and hazard that is driven by these factors rather than underlying host-pathogen associations [@wille_how_2021; @redding_evaluating_2017; @gibb_mammal_2021]. Predictions of zoonotic disease spillover and novel zoonotic pathogen emergence must account for these biases to understand the future hazard of zoonotic diseases [@carlson_future_2021].

West Africa has been identified as a region at increased risk for rodent-borne zoonotic disease spillover events, the probability of these events are predicted to increase under different projected future land-use change scenarios [@grace_mapping_2012; @garcia-pena_land-use_2021]. Currently within West Africa, rodents are involved in the transmission of multiple endemic zoonoses with large burdens on human health, these pathogens include, Lassa fever, Schistosomiasis, Leptospirosis and Toxoplasmosis [@meerburg_rodent-borne_2009; @galeh_global_2020]. Understanding of the distribution of these zoonoses are limited by biases in consolidated datasets. Rodent trapping studies provide contextually rich information on when, where and under what conditions rodents were trapped, potentially enriching consolidated datasets [@bovendorp_optimising_2017]. Studies have been conducted in West Africa to investigate the distribution of rodent species, their species assemblages, the prevalence of endemic zoonoses within rodent hosts (e.g., Lassa fever, Schistosomiasis) and to identify emerging and novel zoonotic pathogens [@fichet-calvet_diversity_2009; @catalano_multihost_2020; @usaid_one_2021]. However, individual level data from these studies have not previously been synthesised for inclusion in assessments of zoonotic disease spillover and novel zoonotic pathogen emergence.

Here, we synthesise rodent trapping studies conducted across West Africa published between 1964-2022. First, we use this dataset to investigate the geographic sampling biases in relation to human population density and land use classification. Second, we compare this to curated host datasets (IUCN and GBIF) to understand differences in reported host geographic distributions. Third, we compare identified host-pathogen associations with a consolidated dataset (CLOVER) to explore discrepancies in rodent host-pathogen associations and report the proportion of positive assays for pathogens of interest. Finally, within our dataset we investigate the spatial extent of current host-pathogen sampling to identify areas of sparse sampling of pathogens within their host ranges. We expect that rodent trapping studies provide an important additional source of high-resolution data that can be used to enrich available consolidated datasets to better understand the hazard of zoonotic disease spillover and novel zoonotic pathogen emergence across West Africa.

@@

We included studies for further analysis if they met all of the following inclusion criteria; i) reported findings from trapping studies where the target was a small mammal, ii) described the type of trap used or the length of trapping activity or the location of the trapping activity, iii) included trapping activity from at least one West African country, iv) recorded the genus or species of trapped individuals, and v) were published in a peer-reviewed journal or as a pre-print on a digital platform or as a report by a credible organisation. We excluded studies if they met any of the following exclusion criteria: i) reported data that were duplicated from a previously included study, ii) no full text available, iii) not available in English. One author screened titles, abstracts and full texts against the inclusion and exclusion criteria. At each stage; title screening, abstract screening and full text review, a random subset (10%) was reviewed by a second author.

We extracted data from eligible studies using a standardised tool that was piloted on 5 studies (Supplementary Table 1.). Data was abstracted into a Google Sheets document, which was archived on completion of data extraction [ref to zenodo on finalised]. We identified the aims of included studies, for example, whether it was conducted as a survey of small mammal species or specifically to assess the risk of zoonotic disease spillover. we extracted data on study methodology, such as, the number of trap nights, the type of traps used and whether the study attempted to estimate abundance. For studies not reporting number of trap nights we used imputation based on the number of trapped individuals, stratified by the habitat type from which they were obtained. This was performed by multiplying the total number of trapped individuals within that study site by the median trap success for study sites with the same reported habitat type. Stratification was used as trap success varied importantly between traps placed in or around buildings (13%, IQR 6-24%) compared with other habitats (3%, IQR 1-9%)

We also recorded how species were identified within a study and species identification was assumed to be accurate. The number of individuals of these species or genera was extracted with taxanomic names mapped to GBIF taxonomy [@gbif_the_global_biodiversity_information_facility_global_2021]. We expanded species detection and non-detection records by explicitly specifying non-detection at a trap site if a species was recorded as detected at other trapping locations within the same study.

Geographic locations of trapping studies were extracted using GPS locations for the most precise location presented. Missing locations were found using the National Geospatial-Intelligence Agency GEOnet Names Server [@national_geospatial-intelligence_agency_nga_nodate] based on place-names and maps presented in the study. All locations were converted to decimal degrees. The year of rodent trapping was extracted alongside the length of the trapping activity to understand seasonal representativeness of trapping activity. The habitats of trapping sites were mapped to the IUCN Habitat Classification Scheme (Version 3.1). For studies reporting multiple habitat types for a single trap, trap-line or trapping grid, a higher order classification of habitat type was recorded.

@@

Out of 4,692 relevant citations, we identified 127 rodent trapping studies (Supplementary Table 2.). The earliest trapping studies were conducted in 1964, with a trend of increasing numbers of studies being performed annually since 2000. The median year of first trapping activity was 2007, with the median length of trapping activity being 1 year (IQR 0-2 years) (Supplementary Figure 1.). Studies were conducted in 14 West African countries, with no studies reported from The Gambia or Togo, at 1,611 trap sites (Figure 1A.).

Included studies explicitly reported on 601,184 trap nights, a further 341,445 trap nights were imputed from studies with no recording of trapping effort based on trap success, leading to an estimate of 942,629 trap nights (Figure 1B.). A minority of studies trapped at a single study site (30, 24%), with 46 (36%) trapping at between two and five sites, the remaining 51 studies (40%) trapped at between six and 93 study sites.

In total 76,275 small mammals were trapped with 65,628 (90%) identified to species level and 7,439 (10%)  identified to genus, with the remaining classified to higher taxonomic level. The majority of the 132 identified species were Rodentia (102, 78%), of which Muridae (73, 72%) were the most common family. Soricomorpha were the second most identified order of small mammals (28, 21%). 57 studies tested for 32 microorganisms, defined to species or genus level that are known or potential pathogens. Most studies tested for a single microorganism (48, 84%). The most frequently assayed microorganisms were *Lassa mammarenavirus* or Arenaviridae (21, 37%), *Borrelia sp.* (9, 16%), *Bartonella sp.* (4, 7%) and *Toxoplasma gondii* (4, 7%). Most studies used Polymerase Chain Reaction (PCR) to detect microorganisms (37, 65%), with fewer studies using serology-based tests (11, 19%) or histological or direct visualisation assays (11, 21%). From 32,920 individual rodent samples we produced 351 host-pathogen pairs. With *Rattus rattus*, *Mus musculus*, *Mastomys erythroleucus*, *Mastomys natalensis* and *Arvicanthis niloticus* being assayed for at least 18 microorganisms.

@@

To investigate the extent of spatial bias in the rodent trapping data, we calculated trap-night (TN) density within each West African level-2 administrative region. The `sf` package in the R statistical language was used to manipulate geographic data, administrative boundaries were obtained from GADM 4.0.4 [@pebesma_simple_2018; @r_core_team_r_2020; @database_of_global_administrative_areas_gadm_2022].  Density ($\text{TN}_\text{density}$) was calculated by dividing the number of trap nights by the area of a level-2 administrative area ($\text{R}_\text{area}$). For studies not reporting trap nights imputation was used as previously described. Human population density was obtained for the closest year (2005) to the median year of trapping (2007) from Socioeconomic Data and Applications Center (SEDAC) gridded population of the world v4 at ~ 1km resolution ($\text{P}_\text{density}$) [@socioeconomic_data_and_applications_center_gridded_2021].  Median population density was then calculated for each level-2 administrative region. Land cover classification was obtained from the Copernicus climate change service at ~300m resolution [@european_space_agency_climate_change_initiative_land_2022]. The proportion of cropland, shrubland, tree cover (${\psi}_\text{tree}$) and urban land cover (${\psi}_\text{urban}$) within a level-2 administrative region in 2005 was calculated.

We investigated the association between relative trapping effort, measured as TN density, and the proportion of urban, cropland, tree cover and human population density using Generalised Additive Models (GAM) incorporating a spatial interaction term (longitude and latitude) (@pedersen_hierarchical_2019). The models were constructed in the `mgcv` package (@wood_generalized_2017). Selection of the most parsimonious model was based on Deviance explained and the Akaike Information Criterion for each model (Equations 1-5 below).  Relative trapping effort was then predicted across West Africa using these covariates. We performed two sensitivity analyses, first, by removing sites with imputed trapping effort. Second rather than associating trap locations to level-2 administrative areas we assigned them to ~ 1km pixels.

$$
\text{TN}_\text{density} \sim \text{Tweedie}(X * Y)
$$
$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + (X * Y))
$$

$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + \text{R}_\text{area} + (X * Y))
$$

$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{P}_\text{density} + {\psi}_\text{tree} + {\psi}_\text{urban} + (X * Y))
$$

$$
\text{TN}_\text{density} \sim \text{Tweedie}(\text{Pop}_\text{density} + \text{R}_\text{area} + {\psi}_\text{urban} + (X * Y))
$$

@@

We assessed the concordance of curated rodent host distributions from IUCN and GBIF with observed rodent presence and absence from rodent trapping studies for seven species with the most trap locations (*M. natalensis*, *R. rattus*, *M. erythroleucus*, *M. musculus*, *A. niloticus*, *Praomys daltoni* and *Cricetomys gambianus*). We obtained rodent species distribution maps as shapefiles from the IUCN red list and translated these to a ~20km resolution raster [@iucn_iucn_2021]. Distributions were cropped to the study region for globally distributed rodent species. We obtained rodent presence locations from GBIF as point data limited to the study region [@gbif_the_global_biodiversity_information_facility_occurrence_2021]. Presence locations were associated to cells of raster with a ~20km resolution produced for the study region.

For each of the seven species, we first calculated the area of the IUCN expected range, and then the percentage of this range covered by presence detections in GBIF, and from presence detections in the rodent trapping data. We then calculated the area of both types of detections outside of the IUCN expected range. For rodent trapping data, we calculated the area of non-detections within the IUCN expected area. Finally, we calculated the combined area of detection from both GBIF and rodent trapping data.

@@

To examine the usefulness of rodent trapping studies as an additional source of data we compared identified host-pathogen associations from trapping studies investigating zoonoses with a consolidated zoonoses dataset (CLOVER) [@gibb_mammal_2021l @gibb_viralemergenceclover_2021]. CLOVER is a synthesis of four host-pathogen datasets (GMPD2, EID2, HP3 and Shaw, 2020) and was released in 2021, it contains more than 25,000 host-pathogen associations for Bacteria, Viruses, Helminth, Protozoa and Fungi. We compared the host-pathogen networks across the two datasets, where the CLOVER data was subset for host species present in the rodent trapping datta.

For host-pathogen pairs with assay results consistent with acute or prior infection, we calculated the proportion positive and identify those absent from CLOVER. We expand the analysis to host-pathogen pairs with pathogens identified to genus level in Supplementary Figure 4.

### What is the spatial extent of pathogen testing within host ranges?

We use the sampled area of three pathogen groups and two pathogens (Arenaviridae, Borreliaceae, Leptospiraceae, *Lassa mammarenavirus* and *Toxoplasma gondii*) to quantify the bias of sampling within their hosts ranges. For each pathogen, we first describe the number of host species assated, for the five most commonly tested species we associate the locations of sampled indivuals to ~20km pixels and calculate the proportion of the IUCN range of the host in which sampling has occurred. We compare this figure to the total area in which the host has been detected to produce a measure of relative completeness of sampling within the included rodent trapping studies.

All data and code to reproduce these analyses are available in an archived Zenodo repository [ref_zenodo].

@@

table_2 <- read_rds(here("tables", "Table_2_updated_pixel.rds"))%>%
  set_caption(caption = "Table 2: Comparison of pathogen sampling ranges for the 5 most widely sampled pathogens and the 5 most sampled rodent host species. * no IUCN range in West African") %>%
  autofit()
